VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-10 doc/publications/journal2014/paper.tex
updated paper

As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
or our Hadoop implementation (14min and 22secs). %862s
Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, the ADAM framework requires the input to be in the ADAM file format, and VCF files must therefore be converted to this binary format. 
Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs, %793s
rendering our approach almost an order of magnitude faster. 
R and Python are slower at pre-processing, taking 67 minutes and 92 minutes, respectively. This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, ADAM and Hadoop can use all eight available cores.

Clustering the samples is fastest in \variantSpark{} (1min and 20secs), being 30\% faster than ADAM (1min and 52secs) and  %80 vs 112
90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib \kMeans{} implementation. 
The 30 second speed increase in \variantSpark{} is likely due to us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.

To demonstrate {\sc Spark}'s advantages over parallel approaches to \kMeans{} clustering in existing libraries, we compare the clustering libraries used in {\variantSpark}, to those in Python and R. For this comparison, each implementation ran on 8 cores.
In regards to speed, clustering using Python and R were both an order of magnitude slower than \variantSpark{} and ADAM, however both were faster than our Hadoop implementation. Python required 11mins and 29secs and R required 7mins and 25secs.
The increased speed compared to Hadoop is due the dataset (chromosome 22) being small enough to fit into memory. This enables Python and R to store the output from each \kMeans{} iteration in memory. Hadoop, however, writes the entire output of each iteration to disk.
Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
This demonstrates one of the major improvements of {\sc Spark} over Hadoop, {\sc Spark}'s in-memory caching of datasets. Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where only a portion of the data can be cached in memory.

For smaller datasets, we've demonstrated {\sc Spark} to be faster than its single-machine counterparts, however it's also important that this speedup isn't at the cost of quality.
The 1000 Genomes data is ideal for investigating the clustering quality, as the consortium provides metadata which lists the super-population that each individual belongs to. This knowledge allows us to create a `truth' clustering, where individuals are clustered by their super-population.
With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. We can then compare this truth clustering with the \kMeans{} clusters.
For this comparison, we use the \ARI{} (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
Two identical clusterings will result in a 1, whereas 0 represents two clusterings that do not agree on any points (individuals). On Chromosome 22, each or the algorithms resulted in an ARI of 0.84. Although not perfect, the identical results demonstrate that the speed-up in {\sc Spark} isn't at the cost of quality.
A result of 1 is probably unlikely because, as mentioned by 10.15252/embr.201439469, although differences in the distribution of alleles exist between populations, the boundaries are rather diffuse. %TODO - proper citation

@@

Although we've demonstrated the advantages of \variantSpark{} with smaller datasets, the main advantage of {\sc Spark} (and hense \variantSpark) is it's scalability and ability to process datasets that exceed memory limits.
To demonstrate the scalability, we continue to cluster the same individuals from the 1000 Genomes Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.
We run this comparison on our in-house Hadoop cluster. Pre-processing the VCF files takes approximately 12 minutes, 19 minutes and 40 minutes for 20\%, 40\% and 100\% of the genome, respectively.
In each case, the memory required doesn't exceed a modest 2GB per executor, even for 100\% of the genome. Memory requirements for clustering, however, increase in correlation with the dataset.

Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, scaling to 100\% of the genome increases the requirement to more than 24GB. This scales in an approximate linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
24GB per executor for \kMeans{} clustering makes it much more resource intensive than pre-processing's 2GB per executor. 
One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants. Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors in our pre-processing stage. However, to cluster these vectors, the \kMeans{} algorithm converts these sparse vectors to dense vectors. This is much less memory-efficient, as dense vectors explicitly define zero values (which are excluded from Sparse Vectors so don't require any memory), considerably increasing the memory required.
So although we demonstrate the potential of \kMeans{} clustering to scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, or for example, easily deal with more alleles.

\subsection*{Personal Genome Project}
To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP). The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features ~\cite{Lunshof2013}.
The  VCF file we use contains 985,790 alleles.

Clustering individuals leads to an ARI of 0.1. 
Some observations are that the population groups to diverse in the PGP data. Each individual has metadata stating which country each of their grandparents are from, and only 40 individuals have all four grandparents from the same country.
Also, being open to the public, the data may be more prone to error. For some individuals, the stated country of their grandparents changes between their updates. This could indicate some uncertainty about their origins, and hence, correct population.

To try improve on these results, we only include the individuals with their grandparents were from the same country.
We then grouped them into their approximate super-populations. Clustering these individuals results in an ARI of 0.29. Although still a far cry from the 1000 Genomes Project results, it's an improvement on the previous clustering, albeit with fewer individuals.

>>>>

As shown in Table~\ref{fivewaycomparison}, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
This is almost 80\% faster than the ADAM implementation (12min and 48secs) %768s
or our Hadoop implementation (14min and 22secs). %862s
Unlike \variantSpark{} and our Hadoop implementation, the ADAM framework cannot process the VCF files directly but requires them to be converted into a binary ADAM file format. 
%Unlike \variantSpark{} or our Hadoop implementation, which pre-process VCF files directly, the ADAM framework requires the input to be in the ADAM file format, and VCF files must therefore be converted to this binary format. 
Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
rendering our approach almost an order of magnitude faster. 
%Although this additional pre-processing step is only required as a one-off for each input file, it requires an additional 13mins and 13secs, %793s
rendering our approach almost an order of magnitude faster. 
R and Python are slower at pre-processing, taking 67 minutes and 92 minutes, respectively. This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, ADAM and Hadoop can use the eight available cores.

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
Out Hadoop implementation performed the worst with 14mins and 23secs.
The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory. This enables Python and R to store the output from each \kMeans{} iteration in memory. Hadoop, however, writes the entire output of each iteration to disk.
Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
{\sc Spark} overcomes this limitation by performing in-memory caching of datasets. Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.

%Clustering the samples is fastest in \variantSpark{} (1min and 20secs), being 30\% faster than ADAM (1min and 52secs) and  %80 vs 112
%90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
%The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib \kMeans{} implementation. 
%The 30 second speed increase in \variantSpark{} is likely due to us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.

%To demonstrate {\sc Spark}'s advantages over parallel approaches to \kMeans{} clustering in existing libraries, we compare the clustering libraries used in {\variantSpark}, to those in Python and R. 
%For this comparison, each implementation ran on 8 cores.
%In regards to speed, clustering using Python and R were both an order of magnitude slower than \variantSpark{} and ADAM, however both were faster than our Hadoop implementation. Python required 11mins and 29secs and R required 7mins and 25secs.
%The substantially increased speed compared to Hadoop is due the dataset (chromosome 22) being small enough to fit into memory. This enables Python and R to store the output from each \kMeans{} iteration in memory. Hadoop, however, writes the entire output of each iteration to disk.
%Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
%This demonstrates one of the major improvements of {\sc Spark} over Hadoop, {\sc Spark}'s in-memory caching of datasets. Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where only a portion of the data can be cached in memory.

We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters. 
For this comparison, we use the \ARI{} (ARI) metric, which returns a value between 1 (perfect) and 0 (random clustering). 
Using Chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
Including AMR individuals, we observe that the majority of these individuals are placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Hunter2014}.

%For smaller datasets, we've demonstrated {\sc Spark} to be faster than its single-machine counterparts, however it's also important that this speedup isn't at the cost of quality.
%The 1000 Genomes data is ideal for investigating the clustering quality, as the consortium provides metadata which lists the super-population that each individual belongs to. This knowledge allows us to create a `truth' clustering, where individuals are clustered by their super-population.
%With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. 
%We can then compare this truth clustering with the \kMeans{} clusters.
%For this comparison, we use the \ARI{} (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
%Two identical clusterings will result in a 1, whereas 0 represents two clusterings that do not agree on any points (individuals). 
%On Chromosome 22, each or the algorithms resulted in an ARI of 0.84. Although not perfect, the identical results demonstrate that the speed-up in {\sc Spark} isn't at the cost of quality.
%A result of 1 is probably unlikely because, as mentioned by Hunter {\it et al.}~\cite{Hunter2014}, although differences in the distribution of alleles exist between populations, the boundaries are rather diffuse.

@@

Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
% TODO add numbers
To demonstrate this scalability, we increase the number of variants used in the clustering from the initial XX on chromosome 22 to YY including all variants in the genome.

%Although we have demonstrated the advantages of \variantSpark{} on smaller datasets, the main advantage of {\sc Spark} (and hense \variantSpark) is it's scalability and ability to process datasets that exceed memory limits.
%To demonstrate the scalability, we continue to cluster the same individuals from the 1000 Genomes Project, however, rather than clustering on a single chromosome, we cluster the individuals on different subsets of variants from their 22 autosomes, and finally every variant.

% TODO add numbers
We run this comparison on our in-house Hadoop cluster using with XX cores.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes and 40 minutes for 20\%, 40\% and 100\% of the genome, respectively.
In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
%TODO add numbers
Clustering the data takes XX minutes, XX minutes and XX minutes, respectively.
For clustering, the memory requirements increases approximate linearly with 6GB, 12GB and 24GB, respectively, per executor. 
%Memory requirements for clustering, however, increase in correlation with the dataset. 
%Where our previous clustering of chromosome 22 required no more than 4GB memory per executor, scaling to 100\% of the genome increases the requirement to more than 24GB. 
%This scales in an approximate linear trend, where 20\% of the genome requires 6GB and 40\% of the genome requires 12GB.
%24GB per executor for \kMeans{} clustering makes it much more resource intensive than pre-processing's 2GB per executor. 
This substantial increase in memory requirement is due to \kMeans{} converting the provided sparse vectors into dense vectors during processing.  
% DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors, 
% TODO: is it farther ?
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?

%One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants. Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors in our pre-processing stage. However, to cluster these vectors, the \kMeans{} algorithm converts these sparse vectors to dense vectors. This is much less memory-efficient, as dense vectors explicitly define zero values (which are excluded from Sparse Vectors so don't require any memory), considerably increasing the memory required.
%So although we demonstrate the potential of \kMeans{} clustering to scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, or for example, easily deal with more alleles.

\subsection*{Personal Genome Project}
To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP). The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features ~\cite{Lunshof2013}.
The VCF file we use contains 985,790 variants.

% TODO add number
Again we cluster the individuals allowing X clusters and compare the assigned labels to the same number of self-reported population labels.
The observed ARI is 0.1. 
This poor result may be due to the population annotation being too detailed and diverse especially so for self-reported and hence potential uncertain ancestry.
%Some observations are that the population groups to diverse in the PGP data. 
%Each individual has metadata stating which country each of their grandparents are from, and only 40 individuals have all four grandparents from the same country.
%Also, being open to the public, the data may be more prone to error. For some individuals, the stated country of their grandparents changes between their updates. This could indicate some uncertainty about their origins, and hence, correct population.
% TODO provide mapping
We therefore only include individuals where all grandparents were reported to be from the same country and group them into their approximate super-populations (see Supplemental Table X). 
This new dataset comprises XX individuals.
%To try improve on these results, we only include the individuals with their grandparents were from the same country.
%We then grouped them into their approximate super-populations. 
Clustering these individuals results in an ARI of 0.29.

%DENIS: If you see an increase in accuracy above we can hypothesise here that 985,790 variants are not sufficient... otherwise a sentence around the population mix in potential real world clinical cohorts....
